Climate indicators

### 2023-03-06 DRAFT_climate_indicators.Rmd
finish cleaning

    -   Aggregate across reference years to get median and sd values (`st_apply)`

@@

```{r, eval=F}
saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.rds")
write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff")
```

@@

```{r}
temp_median_summer <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
                                 proxy=F)
# This is too slow:
# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/temp_median_summer.rds")
# but it could be that RData works better 
```

```{r, eval=T}
names(temp_median_summer)
```

```{r, eval=T}
dim(temp_median_summer)
```

```{r, eval=T}
st_get_dimension_values(temp_median_summer, "Year")
```

@@

```{r}
st_get_dimension_values(temp_median_summer, "band")
```

@@

```{r}
temp_median_summer <- temp_median_summer %>% 
  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
  setNames("temperature")
dim(temp_median_summer)
```

```{r, fig.cap="Showing three random slices of the year dimension."}
ggplot()+
  geom_stars(data = temp_median_summer[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~v_YEAR)
```

@@

We need to first to define a reference period and then to subset our data `temp_median_summer`. We can practice on the dummy data `y`.

@@

```{r}
new_dims <- as.Date(paste0(substr(st_get_dimension_values(temp_median_summer, "v_YEAR"), 3, 6), "-01-01"))
temp_median_summer_ref <- temp_median_summer %>%
  st_set_dimensions("v_YEAR", values = new_dims)
```

@@

```{r}
temp_median_summer_ref <- temp_median_summer_ref %>%
  filter(v_YEAR %within% interval("1961-01-01", "1990-12-31"))
st_get_dimension_values(temp_median_summer_ref, "v_YEAR")
```

@@

```{r, eval=FALSE}
system.time({
cl <- makeCluster(10L)
temp_median_summer_ref <- temp_median_summer_ref %>%
  st_apply(c("x", "y"), 
           FUN =  median_sd,
           CLUSTER = cl)
stopCluster(cl)
})
```

user system elapsed 9.624 6.069 20.903

@@

```{r, eval=F}
# write_stars can only export a single attribute 
#write_stars(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.tiff")

saveRDS(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
```

```{r, eval=T, include=F}
temp_median_summer_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
```

```{r, eval=T}
temp_median_summer_ref_long <- temp_median_summer_ref %>% split("median_sd")
```

@@

```{r, eval=T}
temp_median_summer_ref_long <- setNames(temp_median_summer_ref_long, c("reference_upper", "sd"))
```

```{r, fig.cap="Showing the upper reference levels and the standard deviation from actual data of median summer temperatures."}
tmap_arrange(
tm_shape(st_downsample(temp_median_summer_ref_long, 10))+
  tm_raster("reference_upper")
,
tm_shape(st_downsample(temp_median_summer_ref_long, 10))+
  tm_raster("sd",
            palette = "-viridis")
)
```

@@

```{r}
y_var <- temp_median_summer %>%
  split(3) %>%
  c(temp_median_summer_ref_long)

@@

```{r, eval=F}
# Mutate
# can I find a way to keep the v_ columns?

@@

user system elapsed 14.803 2.717 17.512

```{r, eval=F}
gc()
saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
```

```{r, include=FALSE}
y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
```

@@

The *real* indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming. Therefore, the time series will be presented with yearly resolution, and the scaled, regional indicator values will be calculated for distinct time steps.

@@

We could also plot a time series as a series of maps, but then we need to avergae over this 5 time periods firs. We can get to that later.

@@

However, this does not preserve the region name, and a subsequent st_intersection taks a little whie and gives some boundary problems.

I can also converting att pixels to points and then do the intersection.

@@

I can rastereze the regions data set, turn it into a coded dimension and use st_apply, but thats tedious.

Let me try using `exactextract`, as a parallel to `aggregate`. This requires us to turn to the `terra` package for a bit.

@@

user system elapsed 7.603 3.071 10.697

@@

user system elapsed 7.420 2.921 10.355

@@

```{r}
regOrder = c("Østlandet","Sørlandet","Vestlandet","Midt-Norge","Nord-Norge")

@@

```{r}
myCol <- "RdYlGn"
myCol2 <- "-RdYlGn"

@@

#tm_main <- tmap_grob(tm_main)
#tm_inset <- tmap_grob(tm_inset)

#ggdraw() +
#  draw_plot(tm_main) +
#  draw_plot(tm_inset,
#    height = 0.4,
#    x = 0.12, y = .35
#  )

@@

```{r}
  ts1 <- y_var_norm %>%
  select(all_of(starts_with("i_"))) %>%
  st_redimension() %>%
  st_apply(3, median_sd) %>%
  as_tibble() %>%
  setNames(c("metric", "i_year", "value") ) %>%
  mutate(year = as.numeric(substr(i_year, 3, 7))) %>%
  pivot_wider(names_from = "metric") %>%
  select(-i_year) %>%
  add_column(type = "indicator_value") %>%
  rename(value = median) %>%
  add_column(metric = "median")
```

```{r}
ts2 <- y_var_norm %>%
  select(all_of(starts_with("v_"))) %>%
  st_redimension() %>%
  st_apply(3, median_sd) %>%
  as_tibble() %>%
  setNames(c("metric", "v_year", "value") ) %>%
  mutate(year = as.numeric(substr(v_year, 3, 7))) %>%
  pivot_wider(names_from = "metric") %>%
  select(-v_year) %>%
  add_column(type = "variable_value") %>%
  rename(value = median) %>%
  add_column(metric = "median")
```

```{r}
  ts3 <- y_var_norm %>%
  select(!all_of(starts_with(c("v_", "i_")))) %>%
  st_redimension() %>%
  st_apply(3, median) %>%
  as_tibble() %>%
  setNames(c("type", "value")) %>%
    add_column(metric = "median")
```

```{r}
(ts <- bind_rows(ts1, ts2, ts3))
```

Calculate diference between indicator and reference We don't want to include this in ts, but as a temp file.

```{r}
tsX <- ts1
```

```{r}
ts %>% filter(type == "indicator_value") %>%
  ggplot(aes(x = year, y = value))+
  geom_line()
```

### Scaled indicator values

<!-- Text and analyses here -->

### Uncertainty

<!-- Text here -->

## Prepare export

<!-- Text here -->

### Eksport file (final product)

<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

>>>>

    -   Aggregate (`st_apply)` across reference years (`dplyr::filter`) to get median and sd values

@@

```{r, eval=F}
saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.rds")
write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff")
```

@@

```{r}
summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff",
                                 proxy=F)
# This is too slow:
# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/summer_median_temp.rds")
# but it could be that RData works better 
```

```{r, eval=T}
names(summer_median_temp)
```

```{r, eval=T}
dim(summer_median_temp)
```

```{r, eval=T}
st_get_dimension_values(summer_median_temp, "Year")
```

@@

```{r}
st_get_dimension_values(summer_median_temp, "band")
```

@@

```{r}
summer_median_temp <- summer_median_temp %>% 
  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
  setNames("temperature")
dim(summer_median_temp)
```

```{r, fig.cap="Showing three random slices of the year dimension."}
ggplot()+
  geom_stars(data = summer_median_temp[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~v_YEAR)
```

@@

We need to first to define a reference period and then to subset our data `summer_median_temp`. We can practice on the dummy data `y`.

@@

```{r}
new_dims <- as.Date(paste0(
  substr(st_get_dimension_values(summer_median_temp, "v_YEAR"), 3, 6), "-01-01"))
summer_median_temp_ref <- summer_median_temp %>%
  st_set_dimensions("v_YEAR", values = new_dims)
```

@@

```{r}
summer_median_temp_ref <- summer_median_temp_ref %>%
  filter(v_YEAR %within% interval("1961-01-01", "1990-12-31"))
st_get_dimension_values(summer_median_temp_ref, "v_YEAR")
```

@@

```{r, eval=FALSE}
system.time({
cl <- makeCluster(10L)
summer_median_temp_ref <- summer_median_temp_ref %>%
  st_apply(c("x", "y"), 
           FUN =  median_sd,
           CLUSTER = cl)
stopCluster(cl)
})
```

| user  | system | elapsed |
|-------|--------|---------|
| 9.624 | 6.069  | 20.903  |

@@

```{r, eval=F}
# write_stars can only export a single attribute 
#write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")

saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

```{r, eval=T, include=F}
summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
```

Pivot and turn dimension into attributes:

```{r, eval=T}
summer_median_temp_ref_long <- summer_median_temp_ref %>% split("median_sd")
```

@@

```{r, eval=T}
summer_median_temp_ref_long <- setNames(summer_median_temp_ref_long, c("reference_upper", "sd"))
```

```{r, fig.cap="Showing the upper reference levels and the standard deviation from actual data of median summer temperatures."}
tmap_arrange(
tm_shape(st_downsample(summer_median_temp_ref_long, 10))+
  tm_raster("reference_upper")
,
tm_shape(st_downsample(summer_median_temp_ref_long, 10))+
  tm_raster("sd",
            palette = "-viridis")
)
```

@@

```{r}
y_var <- summer_median_temp %>%
  split(3) %>%
  c(summer_median_temp_ref_long)

@@

```{r, eval=F}
# Mutate

@@

| user   | system | elapsed |
|--------|--------|---------|
| 14.803 | 2.717  | 17.512  |

```{r, eval=F}
gc()
saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
```

```{r, include=FALSE}
y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
```

@@

The *real* indicator values should be means over 5 year periods. Calculating a running mean for all time steps is too time consuming. Therefore, the scaled, regional indicator values will be calculated for distinct time steps. The time series can perhaps best be presented with yearly resolution.

@@

We could also plot a time series as a series of maps, but then we need to average over these 5 year time periods first. We can get to that later.

@@

However, this does no preserve the region name, and a subsequent `st_intersection` taks a little while and gives some boundary problems.

I can also convert pixels to points and then do the intersection.

@@

I can rasterize the regions data set, turn it into a coded dimension and use st_apply, but that is tedious.

Let me try using `exact_extract`, as an alternative to `aggregate`. This requires us to turn to the `terra` package for a bit.

@@

user; system; elapsed: 7.603; 3.071; 10.697

@@

user; system; elapsed: 7.420; 2.921; 10.355

@@

```{r, fig.cap="Times series for median summer temperature centered on the median value during the reference period. The reference period is indicated with a thick horizontal line. Dottet horisontal lines are 2 sd units for the reference period."}
regOrder = c("Østlandet","Sørlandet","Vestlandet","Midt-Norge","Nord-Norge")

@@

```{r, fig.cap="Summer tempreature indicator values for five accounting areas in Norway. SD is the spatial variation."}
myCol <- "RdYlGn"
myCol2 <- "-RdYlGn"

@@

@@
